### 2021-09-28 data-raw/data-preprocessing.Rmd
remove docs from gitignore, add docs

# Reading the data

@@

</details>
The above source code creates a data set `new_data_qnames`.
As shown below, the column names contain information on the job number
(HRP1 = job 1, HRP2 = job 2, ..., HRP5 = job 5) and the year information.

@@

# Demographic data pre-processing
## Tidy the date of birth data

```{r dob-tidy}
dob_tidy <- new_data_qnames %>%
  select(CASEID_1979,
                starts_with("Q1-3_A~")) %>%
  mutate(dob_year = case_when(
                      # if the years recorded in both sets match, take 79 data
                      `Q1-3_A~Y_1979` == `Q1-3_A~Y_1981` ~ `Q1-3_A~Y_1979`,
                      # if the year in the 81 set is missing, take the 79 data
                      is.na(`Q1-3_A~Y_1981`) ~ `Q1-3_A~Y_1979`,
                      # if the sets don't match for dob year, take the 79 data
                      `Q1-3_A~Y_1979` != `Q1-3_A~Y_1981` ~ `Q1-3_A~Y_1979`),
         dob_month = case_when(
                      # if months recorded in both sets match, take 79 data
                      `Q1-3_A~M_1979` == `Q1-3_A~M_1981` ~ `Q1-3_A~M_1979`,
                      # if month in 81 set is missing, take the 79 data
                      is.na(`Q1-3_A~M_1981`) ~ `Q1-3_A~M_1979`,
                      # if sets don't match for dob month, take the 79 data
                      `Q1-3_A~M_1979` != `Q1-3_A~M_1981` ~ `Q1-3_A~M_1979`),
         # flag if there is a conflict between dob recorded in 79 and 81
         dob_conflict = case_when(
                      (`Q1-3_A~M_1979` != `Q1-3_A~M_1981`) & !is.na(`Q1-3_A~M_1981`)
                      ~ TRUE,
                      (`Q1-3_A~Y_1979` != `Q1-3_A~Y_1981`) & !is.na(`Q1-3_A~Y_1981`)
                      ~ TRUE,
                      (`Q1-3_A~Y_1979` == `Q1-3_A~Y_1981`) &
                        (`Q1-3_A~M_1979` == `Q1-3_A~M_1981`) ~ FALSE,
                      is.na(`Q1-3_A~M_1981`) | is.na(`Q1-3_A~Y_1981`) ~ FALSE)) %>%
  select(CASEID_1979,
         dob_month,
         dob_year,
         dob_conflict)
```

@@

```{r demog-tidy}
demog_tidy <- categories_qnames %>%
  select(CASEID_1979,
         SAMPLE_RACE_78SCRN,
         SAMPLE_SEX_1979) %>%
  rename(gender = SAMPLE_SEX_1979,
         race = SAMPLE_RACE_78SCRN)
```

@@

```{r demog-ed}
demog_education <- new_data_qnames %>%
  as_tibble() %>%
  # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
  rename(HGC_2018 = `Q3-4_2018`) %>%
  select(CASEID_1979,
                starts_with("HGCREV"),
                "HGC_2012",
                "HGC_2014",
                "HGC_2016",
                "HGC_2018") %>%
  pivot_longer(!CASEID_1979,
               names_to = "var",
               values_to = "grade") %>%
  separate("var", c("var", "year"), sep = -4) %>%
  filter(!is.na(grade)) %>%
  select(-var)
```

@@

```{r tidy-hgc}
highest_year <- demog_education %>%
  group_by(CASEID_1979) %>%
  mutate(hgc_i = max(grade)) %>%
  filter(hgc_i == grade) %>%
  filter(year == first(year)) %>%
  rename(yr_hgc = year) %>%
  select(CASEID_1979, yr_hgc, hgc_i) %>%
  ungroup() %>%
  mutate(hgc = case_when(hgc_i == 0 ~ "NONE",
                         hgc_i == 1 ~ "1ST GRADE",
                         hgc_i == 2 ~ "2ND GRADE",
                         hgc_i == 3 ~ "3RD GRADE",
                         hgc_i >= 4 & hgc_i <= 12 ~ paste0(hgc_i,"TH GRADE"),
                         hgc_i == 13 ~ "1ST YEAR COL",
                         hgc_i == 14 ~ "2ND YEAR COL",
                         hgc_i == 15 ~ "3RD YEAR COL",
                         hgc_i == 95 ~ "UNGRADED",
                         TRUE ~ paste0((hgc_i - 12), "TH YEAR COL")))

```
```{r full-demog}
full_demographics <- full_join(dob_tidy, demog_tidy, by = "CASEID_1979") %>%
  full_join(highest_year, by = "CASEID_1979") %>%
  rename("id" = "CASEID_1979") %>%
  mutate(age_1979 = 1979 - (dob_year + 1900))

>>>>

# Reading the data {#read-data}

@@

</details>

The above source code creates a data set `new_data_qnames`.
As shown below, the column names contain information on the job number
(HRP1 = job 1, HRP2 = job 2, ..., HRP5 = job 5) and the year information.

@@

# Demographic data pre-processing
## Tidy the date of birth data
The month and year of birth is recorded in 1979 and 1981 for each individual.
The records in 1981 are missing for some individuals so we take the month and year of
birth from 1979 records.

Where the record is present for both 1979 and 1981, we check that the
record matches.

```{r dob-tidy}
dob_tidy <- new_data_qnames %>%
  select(id = CASEID_1979,
         year_1979 = `Q1-3_A~Y_1979`,
         year_1981 = `Q1-3_A~Y_1981`,
         month_1979 = `Q1-3_A~M_1979`,
         month_1981 = `Q1-3_A~M_1981`) %>%
  mutate(dob_year = case_when(
                      # if the years recorded in both sets match, take 79 data
                      year_1979 == year_1981 ~ year_1979,
                      # if the year in the 81 set is missing, take the 79 data
                      is.na(year_1981) ~ year_1979,
                      # if the sets don't match for dob year, take the 79 data
                      year_1979 != year_1981 ~ year_1979),
         dob_month = case_when(
                      # if months recorded in both sets match, take 79 data
                      month_1979 == month_1981 ~ month_1979,
                      # if month in 81 set is missing, take the 79 data
                      is.na(month_1981) ~ month_1979,
                      # if sets don't match for dob month, take the 79 data
                      month_1979 != month_1981 ~ month_1979),
         # flag if there is a conflict between dob recorded in 79 and 81
         dob_conflict = case_when(
                      !is.na(month_1981) && (month_1979 != month_1981) ~ TRUE,
                      !is.na(year_1981) && (year_1979 != year_1981) ~ TRUE,
                      (year_1979 == year_1981) & (month_1979 == month_1981) ~ FALSE,
                      is.na(month_1981) | is.na(year_1981) ~ FALSE)) %>%
  select(id,
         dob_month,
         dob_year,
         dob_conflict)

has_dob_conflict <- any(dob_tidy$dob_conflict, na.rm = TRUE)
```
```{r dob-conflict-no, echo=FALSE, eval=!has_dob_conflict}
cat("All birth month and year recorded in 1979 and 1981 match.")
```
```{r dob-conflict-yes, echo=has_dob_conflict, eval=has_dob_conflict}
cat("The birth record does not match for the following individuals.")
dob_tidy %>%
  filter(dob_conflict)
```

@@

```{r demog-tidy}
demog_tidy <- categories_qnames %>%
  select(id = CASEID_1979,
         race = SAMPLE_RACE_78SCRN,
         gender = SAMPLE_SEX_1979)
```

@@

```{r demog-ed}
demog_education <- new_data_qnames %>%
  # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
  rename(HGC_2018 = `Q3-4_2018`) %>%
  select(id = CASEID_1979,
         starts_with("HGCREV"),
         "HGC_2012",
         "HGC_2014",
         "HGC_2016",
         "HGC_2018") %>%
  pivot_longer(!id,
               names_to = "var",
               values_to = "grade") %>%
  separate("var", c("var", "year"), sep = "_") %>%
  filter(!is.na(grade)) %>%
  select(-var)
```

@@

```{r tidy-hgc}
highest_year <- demog_education %>%
  group_by(id) %>%
  mutate(hgc_i = max(grade)) %>%
  filter(hgc_i == grade) %>%
  filter(year == first(year)) %>%
  rename(yr_hgc = year) %>%
  select(id, yr_hgc, hgc_i) %>%
  ungroup() %>%
  mutate(hgc = case_when(hgc_i == 0 ~ "NONE",
                         hgc_i == 1 ~ "1ST GRADE",
                         hgc_i == 2 ~ "2ND GRADE",
                         hgc_i == 3 ~ "3RD GRADE",
                         hgc_i >= 4 & hgc_i <= 12 ~ paste0(hgc_i, "TH GRADE"),
                         hgc_i == 13 ~ "1ST YEAR COL",
                         hgc_i == 14 ~ "2ND YEAR COL",
                         hgc_i == 15 ~ "3RD YEAR COL",
                         hgc_i == 95 ~ "UNGRADED",
                         TRUE ~ paste0((hgc_i - 12), "TH YEAR COL")))

```
```{r full-demog}
full_demographics <- full_join(dob_tidy, demog_tidy, by = "id") %>%
  full_join(highest_year, by = "id") %>%
  mutate(age_1979 = 1979 - (dob_year + 1900))
